Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-03-14 paper/paper1/ms.tex
writing more prose and editting equations.  looks good

\subsection{Augmenting the cloned model to match data}
Real stars possess two key extrinsic properties.  Rotational broadening $v\sin{i}$ and radial velocity $RV$ depend on the observer's viewing location. The qualifier \emph{extrinsic} distiguishes from stellar \emph{intrinsic} properties, such as $T_{\mathrm{eff}}, \log{g},\mathrm{and\,} [\mathrm{Fe}/\mathrm{H}]$, which appear the same from any viewing location, at least for stars with isotropic surfaces.  The distinction is important because the extrinsic terms act as simple convolutions and translations to the cloned spectrum.  We therefore define an augmented model, which we designate the ``extrinsic'' model, $\mathsf{F}_{ex}$:

\begin{eqnarray}
    \mathsf{S}_{\rm ext}(\bm{\lambda}_S) = \mathsf{S}_{\rm clone}(\bm{v}-RV) * \zeta \left(\frac{\bm{v}}{v\sin{i}}\right) \label{eqn_convolution}
\end{eqnarray}

where $\zeta$ is the convolution kernel for rigid body rotation \citep{2022ApJS..258...31K} and $\bm{v}$ is the spectral axis represented as relative velocity coordinates.

Operationally, the Doppler shift gets applied to the line center positions $\tilde \lambda_{c,j} = (\lambda_{c,j}-\delta \lambda)$, and not the individual wavelength grid point coordinates, $\bm{\lambda}_S$. This choice yields a key benefit: it makes the $RV$ autodiff-aware, meaning that an infinitesimal change to the RV value can easily be sensed through backpropagation by affecting only the line center positions.  The strategy allows relatively low computational and memory cost since it comports with the sparse matrix implementation.

The instrumental resolving power, $R$, acts as a convolution with a Gaussian line profile of width $\sigma=\frac{c}{2.355 R}$.  Real astronomical instruments usually have wavelength-dependent resolving power, which complicates the implementation.  The extent to which this effect matters will depend on the science case.  For now, \texttt{blas\'e} simply assumes a fixed resolving power.

The pixel resampling step could be achieved with two choices.  One would be to simply evaluate the closed-form model at the coordinates of the data pixels $\bm{\lambda_D}$.  This choice would require a closed-form solution to Equation \ref{eqn_convolution}, which has recently been reported \citep{2021arXiv211006271L}.  Instead, we choose the conventional sampled approach, in which we evaluate the extrinsic model at all of the native resolution wavelength coordinates, and then compute the mean value of those pixels within the bounds of each coarse data pixel.  The resampling procedure is autodiff-aware: the same clusters of native resolution coordinates map to the same data pixel coordinates, no matter what the $RV$ is.  The $RV$ only dictates what flux values are realized within those pixels.  The final forward model for \emph{blas\'e} is designated the ``pixel-level'' model to emphasize that we have achieved a model for each datum in the 1D observed spectrum:

\begin{eqnarray}
    \mathsf{F}_{p}(\bm{\lambda}) = \resample{\Big[\mathsf{F}_{ex} * g(R) \Big]} \label{eqn_final_model}
\end{eqnarray}

where $g$ is the Gaussian instrumental convolution kernel and the \texttt{resample[]} operation indicates the average of pixels within the data boundaries.

@@

Equation \ref{eqn_final_model} has all the same conceivably tunable parameters as the cloned model, \emph{plus} $v\sin{i}$ and $RV$.  The resolving power may also be treated as tunable if it is not known or varies with \emph{e.g.} seeing or instrumental configuration.

@@

So instead, we have to apply some amount of \emph{regularization}, a restiction on the allowed values the model parameters can take on.  Fortunately we have a firm theoretical basis to justify this regularization.  We believe our precomputed synthetic spectral models are \emph{quasi-statically correct}: the predicted spectra resemble the unobserved ``latent spectrum'' with lines in the correct place, but just with the wrong amplitudes.  This statement may stem from the fact that it is easier to predict the mere existence of some energy transition of atoms and molecules than it is to predict their transition rates, abundances, temperature and pressure effects, and all the other line strength effects that flow down to how much light a line ultimately absorbs in a stellar atmosphere.

Even still, the regularization step is in some ways the most subjective, since the degree of regularization will control the degree of overfitting or underfitting. The most extreme regularization---the antithesis of the \emph{laissez-faire} scenario---would yield a model too rigid to respond to the data at all, yielding a model entirely unchanged from the cloned PHOENIX model.  So setting the regularization parameter constitutes the only hyperparameter worthy of tuning in \texttt{blas\'e}.  The choice of how to set the regularization is problem-specific.  We default to the following choice. We fix all line parameters except for amplitude, which receives a regularization of 0.01, meaning that its ampltidue should be within about 1\% of its cloned value.  The extrinsic $v\sin{i}$ and $RV$ have no regularization, but in practice they need to be initialized close to their plausible values.

>>>>

\subsection{Augmenting the Stellar Clone Model for $RV$ and $v\sin{i}$ }
Real stars possess two key extrinsic properties.  Rotational broadening $v\sin{i}$ and radial velocity $RV$ depend on the observer's viewing location. The qualifier \emph{extrinsic} distiguishes from stellar \emph{intrinsic} properties, such as $T_{\mathrm{eff}}, \log{g},\mathrm{and\,} [\mathrm{Fe}/\mathrm{H}]$, which appear the same from any viewing location, at least for stars with isotropic surfaces.  The distinction is important because the extrinsic terms act as simple convolutions and translations to the cloned spectrum.  We therefore define an augmented model, which we designate the ``extrinsic'' model, $\mathsf{S}_{\rm ext}$:

\begin{eqnarray}
    \mathsf{S}_{\rm ext}(\bm{\lambda}_S) = \mathsf{S}_{\rm clone}(\lambda_\mathrm{c} - \frac{RV}{c}\lambda_\mathrm{c}) * \zeta \left(\frac{\bm{v}}{v\sin{i}}\right) \label{eqn_convolution}
\end{eqnarray}

where $\zeta$ is the convolution kernel for rigid body rotation \citep{2022ApJS..258...31K}, $\bm{v}$ is the spectral axis represented as relative velocity coordinates, and $*$ denotes the convolution operator.

Operationally, the Doppler shift gets applied to the line center positions and not the individual wavelength grid point coordinates, $\bm{\lambda}_S$. This choice yields a key benefit: it makes the $RV$ autodiff-aware, meaning that an infinitesimal change to the RV value can easily be sensed through backpropagation by affecting only the line center positions.  The strategy allows relatively low computational and memory cost since it comports with the sparse matrix implementation.

\subsection{Joint Stellar and Telluric Model}
The telluric model and stellar models possess distinct pixel sampling and coordinates.  The design of \texttt{blas\'e} permits the telluric model to be re\"evaluated at any wavelength coordinate vector, provided that it adequately samples the underlying lines.  We choose to evaluate the telluric model at the native resolution of the stellar model.  We then simply multiply the stellar model by the tunable telluric transmission:

\begin{eqnarray}
    \mathsf{M}_{\rm joint} = \mathsf{S}_{\rm ext}(\bm{\lambda}_S) \odot \mathsf{T}_{\rm clone}(\bm{\lambda}_S)
\end{eqnarray}

It is only at this stage that we may apply the instrumental broadening kernel.  The instrumental resolving power, $R$, acts as a convolution with a Gaussian line profile of width $\sigma=\frac{c}{2.355 R}$.  Real astronomical instruments usually have wavelength-dependent resolving power, which complicates the implementation.  The extent to which this effect matters will depend on the science case.  For now, \texttt{blas\'e} simply assumes a fixed resolving power.

The data-pixel sampling is much coarser than the model pixel sampling.  This resampling step could be achieved with two choices.  One would be to simply evaluate the closed-form model at the coordinates of the data pixels $\bm{\lambda_D}$.  This choice would require a closed-form solution to Equation \ref{eqn_convolution}, which has recently been reported \citep{2021arXiv211006271L}.  Instead, we choose the conventional sampled approach, in which we evaluate the extrinsic model at all of the native resolution wavelength coordinates, and then compute the mean value of those pixels within the bounds of each coarse data pixel.  The resampling procedure is autodiff-aware: the same clusters of native resolution coordinates map to the same data pixel coordinates, no matter what the $RV$ is.  The $RV$ only dictates what flux values are realized within those pixels.  The final forward model for \emph{blas\'e} is designated simply as $\mathsf{M}$ without subscripts to emphasize that we have achieved the desired goal of a plausible model for each datum in the 1D observed spectrum:

\begin{eqnarray}
    \mathsf{M}(\bm{\lambda}_D) = \resample{\Big[\mathsf{M}_{\rm joint}(\bm{\lambda}_S) * g(R) \Big]} \label{eqn_final_model}
\end{eqnarray}

where $g$ is the Gaussian instrumental convolution kernel and the \texttt{resample[]} operation indicates the average of model pixels that fall within each data pixel's red and blue boundaries.

@@

Equation \ref{eqn_final_model} has up to $N_{\mathrm{lines}}\times3 \sim20,000$ tunabale parameters from the star, $N_{\mathrm{lines}}\times3 \sim15,000$ tunabale parameters from the Earth's atmosphere, plus $v\sin{i}$ and $RV$.  That adds up to 35,002 model parameters.  The resolving power may also be treated as tunable if it is not known or varies slightly with \emph{e.g.} seeing, slit-or-fiber illumination, or instrumental configuration.

@@

We apply some amount of \emph{regularization}, a restiction on the allowed values the model parameters can take on.  Fortunately we have a firm theoretical basis to justify this regularization.  We believe our precomputed synthetic spectral models are \emph{quasi-statically correct}: the predicted spectra resemble the unobserved ``latent spectrum'' with lines in the correct place, but just with the wrong amplitudes.  This statement may stem from the fact that it is easier to predict the mere existence of some energy transition of atoms and molecules than it is to predict their transition rates, abundances, temperature and pressure effects, and all the other line strength effects that flow down to how much light a line ultimately absorbs in a stellar atmosphere.

Even still, the regularization step is in some ways the most subjective, since the degree of regularization will control the degree of overfitting or underfitting. The most extreme regularization---the antithesis of the \emph{laissez-faire} scenario---would yield a model too rigid to respond to the data at all, yielding a model entirely unchanged from the cloned PHOENIX and \texttt{TelFit} models.  So regularization constitutes the only hyperparameters worthy of tuning in \texttt{blas\'e}.  The choice of how to set the regularization is problem-specific.  We default to the following choice. We fix all line parameters except for amplitude, which receives a regularization of 0.01, meaning that its ampltidue should be within about 1\% of its cloned value.  The extrinsic $v\sin{i}$ and $RV$ have no regularization, but in practice they need to be initialized close to their plausible values.
